---
title: Early exposure to language (in any modality) supports number concept development - Insights from deaf children acquiring signed and spoken language
author:
- Emily Carrigan, Ph.D. (code author)
- Anna Shusterman, Ph.D.
- Marie Coppola, Ph.D
output:
  html_document:
    toc: yes
    toc_float: yes
  pdf_document:
    toc: yes
    toc_float: yes
  word_document:
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Code last edited February 8 2022 (as part of response to Reviewers)

#Load required packages
```{r echo=TRUE, results='hide'}
library(tidyverse) 
library(ggplot2)
library(colorspace)
library(stargazer)
library(MASS) # for polr function (ordinal logistic regression)
library(generalhoslem) # for testing model fit (lipsitz test and two others)
library(sure) # package for calculating residuals for ordinal logistic regression (https://journal.r-project.org/archive/2018/RJ-2018-004/RJ-2018-004.pdf)
library(sjmisc) # for easy recoding of variables
library(kableExtra)
library(lmtest)
library(broom)
library(beeswarm)
library(ggbeeswarm)
library(ggpubr)
library(table1)
library(VGAM)
library(forcats)
```

# Data Preparation

## Set working directory (this should be wherever your file is saved) & Import data file
```{r echo=FALSE, results='hide'}
#setwd("") 
#setwd("/Users/emilycarrigan/Dropbox/Data Analysis Work/")
GiveN <- read.csv("GN_Smsummary_Lgall_KW_201106.csv", na.strings = c("N/A", "I/R", "", "TBD", "N/R"))
View(GiveN)
```

## Subset data
```{r results='hide'}
GiveN_subset <- subset(GiveN, GiveN$Including.in.Study.=='Yes' & GiveN$Including_GiveN.=='Yes' & GiveN$GN_All_Ceiling_Conservative!='' & GiveN$SES..8.66.!='') # some participant IDs in data file never tested, not tested on this task, or missing demographic data
View(GiveN_subset)
str(GiveN_subset) # check the data
summary(GiveN_subset) # check missing values for  variables
```


## Organize/create relevant variables
```{r}
#ORDER Language Groups the way I want them
GiveN_subset$LanguageGroup <- as.factor(factor(as.character(GiveN_subset$Group_4cat), levels = c("English Early", "ASL Early", "English Later", "ASL Later"), exclude=NA))

##Create Modality and Timing variables and add to dataframe
GiveN_subset$Language_Modality <- factor(ifelse(GiveN_subset$Group_4cat == "English Early" | GiveN_subset$Group_4cat == "English Later", "English", "ASL"), levels = c("English", "ASL"))

GiveN_subset$Language_Timing <- factor(as.character(GiveN_subset$Group_2cat), levels = c("Early", "Later"), exclude="") 

#CHECK whether any participants' "Language Tested" for a particular task does not match their Language_Modality group
GiveN_subset$Language_Match <- ifelse(as.character(GiveN_subset$Child_LanguageTested) == as.character(GiveN_subset$Language_Modality), "Match", "NOT MATCH")

#Convert Age of Language Exposure (currently in months) into years
GiveN_subset$AgeofLangExp_years <- GiveN_subset$Age.of.Exposure..mo./12

#Recode Race to have fewer categories
GiveN_subset$Race_recoded <- dplyr::recode(as.character(GiveN_subset$Race), 'Asian' = "Asian", 'Black or African American'="Black or African American", 'Mixed'="More than one", 'Other'="Not Listed/Missing", 'Unsure, or prefer not to answer' = "Not Listed/Missing", 'White'="White", .missing="Not Listed/Missing")

#Recode Ethnicity to have fewer categories
GiveN_subset$Ethnicity_recoded <- dplyr::recode(as.character(GiveN_subset$Ethnicity), 'Hispanic or Latino' = "Hispanic or Latino", 'Not Hispanic or Latino'="Not Hispanic or Latino", 'Prefer not to answer'="Prefer not to answer/Missing", .missing="Prefer not to answer/Missing")

#CREATE BINARY Knower-level variable from ordinal measure in data file
GiveN_subset$GN_sm_2cat <- factor(ifelse(GiveN_subset$GN_Small_Ceiling._FinalSet == 6, "CP-knower", "NOT CP-knower"), levels = c("NOT CP-knower", "CP-knower"))

##CREATE PERCENT correct Give-N LARGE measure (second analysis) - drawn from data in 5 separate columns 
GiveN_subset <-  dplyr::mutate(GiveN_subset, t7_corr = ifelse(is.na(GiveN_subset$Target7_FinalSet_Correct.), GiveN_subset$Target7_1st_Response_Fish_Correct., GiveN_subset$Target7_FinalSet_Correct.))
GiveN_subset <-  dplyr::mutate(GiveN_subset, t9_corr = ifelse(is.na(GiveN_subset$Target9_FinalSet_Correct.), GiveN_subset$Target9_1st_Response_Fish_Correct., GiveN_subset$Target9_FinalSet_Correct.))
GiveN_subset <-  dplyr::mutate(GiveN_subset, t10_corr = ifelse(is.na(GiveN_subset$Target10_FinalSet_Correct.), GiveN_subset$Target10_1st_Response_Fish_Correct., GiveN_subset$Target10_FinalSet_Correct.))
GiveN_subset <-  dplyr::mutate(GiveN_subset, t12_corr = ifelse(is.na(GiveN_subset$Target12_FinalSet_Correct.), GiveN_subset$Target12_1st_Response_Fish_Correct., GiveN_subset$Target12_FinalSet_Correct.))
GiveN_subset <-  dplyr::mutate(GiveN_subset, t16_corr = ifelse(is.na(GiveN_subset$Target16_FinalSet_Correct.), GiveN_subset$Target16_1st_Response_Fish_Correct., GiveN_subset$Target16_FinalSet_Correct.))

#Calculate mean proportion correct for Give-N trials 7, 9, 10, 12, & 16 (only for kids who got all five trials)

# Code below will appropriately *not* calculate mean (puts "NA") if any trial from above columns has NA value--THIS IS CORRECT, bc doing a proportion correct out of different numbers of trials for different kids is a no-go
GiveN_subset <-  GiveN_subset %>% rowwise() %>% dplyr::mutate(GN_Lg_propcorr_NA=mean(c(t7_corr, t9_corr, t10_corr, t12_corr, t16_corr)))
```

# Basic Demographics, Figures, and Tables

## Demographics Summary Tables
*This block and next block used for generating info used in Table 1 and Table 3
```{r}
##Participant Demographics by Language Group
table1::label(GiveN_subset$Age) <- "Age (years)"
table1::label(GiveN_subset$Sex) <- "Sex"
table1::label(GiveN_subset$Race_recoded) <- "Race"
table1::label(GiveN_subset$Ethnicity_recoded) <- "Ethnicity"
table1::label(GiveN_subset$SES..8.66.) <- "SES"
table1::label(GiveN_subset$AgeofLangExp_years) <- "Age of Language Exposure (yrs)"

##Demographics (four groups)
table1::table1(~Age + Sex + Race_recoded + Ethnicity_recoded + SES..8.66. + AgeofLangExp_years | LanguageGroup, data = GiveN_subset, overall=F)

##Demographics by Timing
table1::table1(~Age + Sex + Race_recoded + Ethnicity_recoded + SES..8.66. + AgeofLangExp_years | Language_Timing, data = GiveN_subset, overall=F)

##Demographics by Modality
table1::table1(~Age + Sex + Race_recoded + Ethnicity_recoded + SES..8.66. + AgeofLangExp_years | Language_Modality, data = GiveN_subset, overall=F)
```

## Testing whether SES or Age differ by Language Modality or Language Timing
```{r}
as.numeric(GiveN_subset$Language_Timing)
with(GiveN_subset, cor.test(SES..8.66., as.numeric(Language_Timing)))
with(GiveN_subset, wilcox.test(SES..8.66. ~ Language_Timing))

with(GiveN_subset, cor.test(Age, as.numeric(Language_Timing)))
with(GiveN_subset, wilcox.test(Age ~ Language_Timing))

as.numeric(GiveN_subset$Language_Modality)
with(GiveN_subset, cor.test(SES..8.66., as.numeric(Language_Modality)))
with(GiveN_subset, wilcox.test(SES..8.66. ~ Language_Modality))

with(GiveN_subset, cor.test(Age, as.numeric(Language_Modality)))
with(GiveN_subset, wilcox.test(Age ~ Language_Modality))
```



## Count & proportion of kids who are CP-Knowers and Non-CP-knowers, by Language Timing & Modality
```{r}
summary(GiveN_subset$Language_Timing)
table(GiveN_subset$GN_sm_2cat, GiveN_subset$Language_Timing, useNA = "always")
prop.table(table(GiveN_subset$GN_sm_2cat, GiveN_subset$Language_Timing), margin=2) #proportions calculated summing across columns

summary(GiveN_subset$Language_Modality)
table(GiveN_subset$GN_sm_2cat, GiveN_subset$Language_Modality, useNA = "always")
prop.table(table(GiveN_subset$GN_sm_2cat, GiveN_subset$Language_Modality), margin=2)#proportions calculated summing across columns
```


## Figure 2. Beeswarn Boxplot looking at CP-Knower vs. Non-CP-Knower w/dot colors by Language Timing & Language Modality

```{r}
jitter <- position_jitter(width=0, height=0.075) # Making overlapping points more visible

#N in code below sums rows from table() functions output in previous block
a <- ggplot(data=GiveN_subset, mapping = aes(x=Language_Timing, y=Age)) + geom_boxplot() + geom_dotplot(mapping=aes(fill=GN_sm_2cat, color=GN_sm_2cat), method = "dotdensity", binaxis = "y", position = jitter, stackdir = "center", dotsize=0.5, stackratio = 1.25) + theme(text = element_text(size=11))  + xlab("Language Timing") + ylab("Age (Years)") +  scale_fill_manual(name = "CP-Knower Status", values=c("tan2", "springgreen3"), labels=c("Non-CP-Knowers (50)", "CP-Knowers (126)")) + scale_color_manual(name = "CP-Knower Status", values=c("tan2", "springgreen3"), labels=c("Non-CP-Knowers (50)", "CP-Knowers (126)"))


#N for code below sums rows from table() functions output in previous block
b <- ggplot(data=GiveN_subset, mapping = aes(x=Language_Modality, y=Age)) + geom_boxplot() + geom_dotplot(mapping=aes(fill=GN_sm_2cat, color=GN_sm_2cat), binaxis = "y", position = jitter, stackdir = "center", dotsize=0.5, stackratio = 1.25) + theme(text = element_text(size=11))  + xlab("Language Modality") + ylab("Age (Years)") +    scale_fill_manual(name = "CP-Knower Status", values=c("tan2", "springgreen3"), labels=c("Non-CP-Knowers (50)", "CP-Knowers (126)")) + scale_color_manual(name = "CP-Knower Status", values=c("tan2", "springgreen3"), labels=c("Non-CP-Knowers (50)", "CP-Knowers (126)"))
ggarrange(a, b, ncol=2, nrow=1, common.legend = TRUE, labels="auto")

```

## Looking at only Early Language groups: graphs & Chi-square comparison
```{r}
#Create dataframe with only early-exposed
GN_early <- subset(GiveN_subset, GiveN_subset$Language_Timing=="Early")

#table of CP-knower status vs group (modality)
table(GN_early$GN_sm_2cat, GN_early$Language_Modality)
prop.table(table(GN_early$GN_sm_2cat, GN_early$Language_Modality), margin=2) #proportions calculated summing across columns
```
<br>

*Chi-square test indicates that CP-Knower achievement does not differ by Language Modality* 


## Figure 3. Ages of CP-Knower Achievement by Language Modality for Early Language Groups
```{r}
s#graph of CP-knower status by Age & Language Modality (Early-exposed only)
ggplot(data=GN_early, mapping = aes(x=Language_Modality, y=Age)) + geom_boxplot() + geom_dotplot(mapping=aes(fill=GN_sm_2cat, color=GN_sm_2cat), binaxis = "y", position = jitter, stackdir = "center", dotsize=0.5, stackratio = 1.25) + theme(text = element_text(size=11), legend.position = c(0.14, 0.87))  + xlab("Early Language Group") + ylab("Age (Years)") +    scale_fill_manual(name = "CP-Knower Status", values=c("tan2", "springgreen3"), labels=c("Non-CP-Knowers (23)", "CP-Knowers (70)")) + scale_color_manual(name = "CP-Knower Status", values=c("tan2", "springgreen3"), labels=c("Non-CP-Knowers (23)", "CP-Knowers (70)"))

chisq.test(table(GN_early$GN_sm_2cat, GN_early$Language_Modality))
```



## Figure 4.  Scatterplots of proportion correct values on Give-N Quantities 7, 9, 10, 12, 16 by Age (years), separated by Language Timing & Language Modality
```{r}
# Create the data frame that will be used for Give-N Large analyses 
# This removes missing ("NA") values, and removes participants who were tested on Give-N higher quantities but who were not CP-knowers as determined by their performance on Give-N quantities 1-6
GN_lg_noNA <- subset(GiveN_subset, GiveN_subset$GN_Lg_propcorr_NA!='' & GiveN_subset$GN_sm_2cat!="NOT CP-knower")

# Turn the proportion correct variable into a factor so it can be used in the ordinal model
GN_lg_noNA$GN_lg_factor <- as.factor(as.character(GN_lg_noNA$GN_Lg_propcorr_NA))


summary(GN_lg_noNA$Language_Timing)
table(GN_lg_noNA$Language_Timing, GN_lg_noNA$GN_Lg_propcorr_NA, useNA = "always")

#N for code below is: summary() values - NA values from table()
GNlg_Timing <- ggplot(data=GN_lg_noNA, aes(x=Age, y=GN_Lg_propcorr_NA, shape=Language_Timing, color=Language_Timing)) + geom_jitter(na.rm=TRUE) + geom_smooth(method=lm, se = FALSE) + scale_y_continuous(breaks = seq(0, 1, by = 0.2), limits=c(0,1)) + labs(x="Age at Test (Years)", y="Give-N Large Proportion Correct") + theme(plot.title = element_text(hjust = 0.5), legend.position = c(0.82, 0.15)) + scale_shape_manual(name="Language\nTiming", values=c(17, 16), labels=c("Early (68)", "Later (48)")) + scale_color_manual(name="Language\nTiming", values=c("deepskyblue3", "firebrick3"), labels=c("Early (68)", "Later (48)"))

summary(GN_lg_noNA$Language_Modality)
table(GN_lg_noNA$Language_Modality, GN_lg_noNA$GN_Lg_propcorr_NA, useNA = "always")

#N for code below is: summary() values - NA values from table()
GNlg_Modality <- ggplot(data=GN_lg_noNA, aes(x=Age, y=GN_Lg_propcorr_NA, shape=Language_Modality, color=Language_Modality)) + geom_jitter(na.rm=TRUE) + geom_smooth(method=lm, se = FALSE) +  scale_y_continuous(breaks = seq(0, 1, by = 0.2), limits=c(0,1)) + labs(x="Age at Test (Years)", y="Give-N Large Proportion Correct") + theme(plot.title = element_text(hjust = 0.5), legend.position = c(0.81, 0.15)) + scale_color_manual(name = "Language\nModality", values=c("slategray4", "purple3"), labels=c("English (73)", "ASL (43)")) + scale_shape_manual(name = "Language\nModality", values=c(19, 15),  labels=c("English (73)", "ASL (43)"))

ggarrange(GNlg_Timing, GNlg_Modality, ncol=2, nrow=1, common.legend = FALSE, labels="auto")

```







# THE MODELS 

### Logistic Regression Predicting CP-Knower Status (CP-knower vs. Non-CP-knower) using data from Give-N Quantities 1-6 (Table 2)
```{r}
null <-glm(formula = GN_sm_2cat ~ 1, data = GiveN_subset, family = "binomial")
summary(null) #resid dev: 210.07   AIC 212.07
j <- glm(formula = GN_sm_2cat ~ SES..8.66. + Language_Modality + Age + Language_Timing, data = GiveN_subset, family = "binomial")
summary(j) # Age & Timing sig., Modality & SES not.

exp(coef(j)) #ODDS RATIOS
	#Kids in the early exposed group are 2.23 times more likely to have achieved CP-knower status than kids in the later group
```

## Logistic Regression fit test
```{r}
with(j, null.deviance - deviance)
with(j, df.null - df.residual)
with(j, pchisq(null.deviance - deviance, df.null - df.residual, lower.tail = FALSE))
#Model fits sig. better than a null model, p<.001
```

## Testing other possible models
```{r}
# Checking for an interaction between Age and Language Timing
k <- glm(formula = GN_sm_2cat ~ SES..8.66. + Language_Modality + Age*Language_Timing, data = GiveN_subset, family = "binomial")
summary(k) #resid dev: 159.75   AIC 171.75  # Age only significant predictor
lmtest::lrtest(j, k) # model with Age * Timing interaction not significantly different from model without the interaction

h <- glm(formula = GN_sm_2cat ~ Language_Modality + Age + Language_Timing, data = GiveN_subset, family = "binomial") #model without SES
summary(h)  # residual deviance: 165.98  AIC: 173.98
lmtest::lrtest(j, h) # model without SES not significantly different from model with SES

p <- glm(formula = GN_sm_2cat ~ SES..8.66. + Age + Language_Timing, data = GiveN_subset, family = "binomial") #model without Language Modality
summary(h)  # residual deviance: 165.98  AIC: 173.98
lmtest::lrtest(j, p) # model without Language Modality not significantly different from model with Language Modality

w <- glm(formula = GN_sm_2cat ~ SES..8.66.+ Language_Modality + Age, data = GiveN_subset, family = "binomial") #model without Language Timing
summary(w) # residual deviance: 166.75  AIC: 174.75
lmtest::lrtest(j, w) # model without Language Timing IS significantly different from model with Language Timing; Looking at AIC and residual deviance, the model WITH Language Timing fits the data better

```


## Logistic Regression model diagnostics (for best-fitting model)

```{r}
# from http://www.sthda.com/english/articles/36-classification-methods-essentials/148-logistic-regression-assumptions-and-diagnostics-in-r/

#Step 1: checking that continuous predictors ARE in fact linearly associated with logit of the outcome

# Select only numeric predictors
continuous_pred <- GiveN_subset %>%
  dplyr::select(c(Age, SES..8.66.)) 
predictors <- colnames(continuous_pred)

#get the predicted probabilities for EACH KID in the whole model (based on their specific values of each predictor)
continuous_pred$probabilities <- predict(j, newdata=GiveN_subset, type = "response")

# Bind the logit and tidy the data for plot
#mutated_probs <- mutate(logit = logitlink(probabilities))

continuous_pred %>%
  mutate(logit = log(probabilities/1-probabilities)) %>%
  gather(key = "predictors", value = "predictor.value", -logit)

#scatterplots!
ggplot(continuous_pred, aes(logit, predictor.value))+
  geom_point(size = 0.5, alpha = 0.5) +
  geom_smooth(method = "loess") + 
  theme_bw() + 
  facet_wrap(~predictors, scales = "free_y")



# Step 2: checking for "INFLUENTIAL VALUES" - extreme data points
plot(j, which = 4, id.n = 3)

# Extract model results
j.data <- augment(j) %>% 
  mutate(index = 1:n()) 

j.data %>% top_n(3, .cooksd)

ggplot(j.data, aes(index, .std.resid)) + 
  geom_point(aes(color = GN_sm_2cat), alpha = .5) +
  theme_bw()

j.data %>% 
  filter(abs(.std.resid) > 3)
#NO INFLUENTIAL OBSERVATIONS IN DATA


##Step 3: MULTICOLLINEARITY?
car::vif(j)  ##None! (bc values not >5 or 10)


# WOHOOO WE HAVE A GOOD MODEL, FOLKS!

#output model
stargazer(j, title = "Logistic Regression Results", align=TRUE, dep.var.labels=c("CP-Knower Status"), covariate.labels=c("Socioeconomic Status (SES)", "Language Modality (ASL)", "Age (Years)", "Timing of Language Exposure (Early)"), single.row=TRUE, keep.stat=c("n", "aic"),  out= "GN_CPknowerbinary_SES_Modality_Age_Timing_220208.htm")
```





## Ordinal Logistic Regression for Give-N quantities 7, 9, 10, 12, 16
Both linear and tobit models had poor measures of model fit and poor residual plots, so using an ordinal model with proportion correct (0, 0.2, 0.4, 0.6, 0.8, or 1.0) as the dependent variable

```{r}
# Turn the proportion correct variable into a factor so it can be used in the ordinal model (same code as above)
GN_lg_noNA$GN_lg_factor <- as.factor(as.character(GN_lg_noNA$GN_Lg_propcorr_NA))

# Base model
GN_lg_null <- polr(formula = GN_lg_factor ~ 1, data = GN_lg_noNA,  method = "probit", Hess = TRUE) 
summary(GN_lg_null) #res dev. 305  AIC: 315

# Model with predictors
GN_lg_polr <- polr(formula = GN_lg_factor ~ SES..8.66. + Language_Modality + Age + Language_Timing, data = GN_lg_noNA,  method = "probit", Hess = TRUE) 
summary(GN_lg_polr)  # res dev 265   AIC: 283

tidy(GN_lg_polr, conf.int = TRUE, p.value=TRUE)
dropterm(GN_lg_polr)

lipsitz.test(GN_lg_polr) #ns
logitgof(GN_lg_noNA$GN_lg_factor, fitted(GN_lg_polr)) #ns
pulkrob.chisq(GN_lg_polr, c("Language_Modality", "Language_Timing")) #ns
```



## checking RESIDUALS
```{r}
sres <- resids(GN_lg_polr)

p1 <- autoplot(sres, what = "qq", distribution = qnorm)
p2 <- autoplot(sres, what = "covariate", x = GN_lg_noNA$SES..8.66., xlab = "SES") #bit of bunching at higher values but we know SES is skewed (also not variable of interest in this model)
p3 <- autoplot(sres, what = "covariate", x = GN_lg_noNA$Language_Modality, xlab = "Language Modality") #good
p4 <- autoplot(sres, what = "covariate", x = GN_lg_noNA$Age, xlab = "Age") #good
p5 <- autoplot(sres, what = "covariate", x = GN_lg_noNA$Language_Timing, xlab = "Language Timing") #good
p6 <- autoplot(sres, what = "covariate", x = GN_lg_noNA$GN_lg_factor, xlab = "Large Number Knowledge") # shows that surrogate residuals underestimated for lower values of Give-N large prop corr (0, 0.2, 0.4, 0.6) and overestimated for  (not sure how much to make of this)
grid.arrange(p1, p2,p3, p4, p5, p6, ncol = 3)  # code that produces Figure 2 in Greenwell et al ppr


## Good fitting model!

stargazer(GN_lg_polr, title = "Ordinal Logistic Regression Results", align=TRUE, dep.var.labels=c("Give-N Large Proportion Correct"), covariate.labels=c("Socioeconomic Status (SES)", "Language Modality (ASL)", "Age (Years)", "Timing of Language Exposure (Later)"), single.row=TRUE, keep.stat=c("n", "aic"),  out= "GN_lgordinal_SES_Modality_Age_Timing_probit_220208.htm")
```


## Code for Tables/Figures in Supplementary materials
```{r}
summary(GN_lg_noNA$LanguageGroup)
table(GN_lg_noNA$LanguageGroup, GN_lg_noNA$GN_Lg_propcorr_NA, useNA = "always")

#N for code below is: summary() values - NA values from table()
GNlg_4Grp <- ggplot(data=GN_lg_noNA, aes(x=GN_lg_factor, y=Age, color=LanguageGroup)) + geom_violin(na.rm=TRUE) + labs(x="Age at Test (Years)", y="Give-N Large Proportion Correct") + theme(plot.title = element_text(hjust = 0.5), legend.position = c(0.15, 0.81))
GNlg_4Grp
```
