---
title: Early exposure to language (in any modality) supports number concept development - Insights from deaf children acquiring signed and spoken language
author:
- Emily Carrigan, Ph.D. (code author)
- Anna Shusterman, Ph.D.
- Marie Coppola, Ph.D
output:
  html_document:
    toc: yes
    toc_float: yes
  pdf_document:
    toc: yes
    toc_float: yes
  word_document:
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Code last edited Nov 30 2021 (response to Reviewers)

```{r}
library(tidyverse) 
library(ggplot2)
library(colorspace)
library(stargazer)
library(MASS) # for polr function (ordinal logistic regression)
library(generalhoslem) # for testing model fit (lipsitz test and two others)
library(sure) # package for calculating residuals for ordinal logistic regression (https://journal.r-project.org/archive/2018/RJ-2018-004/RJ-2018-004.pdf)
library(sjmisc) # for easy recoding of variables
library(broom)
library(beeswarm)
library(VGAM)
```




## Import File & View Summary Characteristics
```{r}
#setwd("") ##this should be wherever your file is saved 

GiveN <- read.csv("GN_Smsummary_Lgall_200304.csv", na.strings = c("N/A", "I/R", "", "TBD", "N/R"))
View(GiveN)

GiveN_subset <- subset(GiveN, GiveN$Including.in.Study.=='Yes' & GiveN$Including_GiveN.=='Yes' & GiveN$GN_All_Ceiling_Conservative!='' & GiveN$SES..8.66.!='') # some participant IDs in data file never tested, not tested on this task, or missing demographic data
View(GiveN_subset)

str(GiveN_subset) # check the data
summary(GiveN_subset) # check missing values for  variables

```


## Data Cleaning/Organization

```{r}
#ORDER Language Groups the way I want them
GiveN_subset$LanguageGroup <- as.factor(factor(as.character(GiveN_subset$Group_4cat), levels = c("English Early", "ASL Early", "English Later", "ASL Later"), exclude=NA))
##Create Modality and Timing variables and add to dataframe
GiveN_subset$Language_Modality <- factor(ifelse(GiveN_subset$Group_4cat == "English Early" | GiveN_subset$Group_4cat == "English Later", "English", "ASL"), levels = c("English", "ASL"))

GiveN_subset$Language_Timing <- factor(as.character(GiveN_subset$Group_2cat), levels = c("Later", "Early"), exclude="") 

#CHECK whether any participants' "Language Tested" for a particular task does not match their Language_Modality group
GiveN_subset$Language_Match <- ifelse(as.character(GiveN_subset$Child_LanguageTested) == as.character(GiveN_subset$Language_Modality), "Match", "NOT MATCH")

#CREATE BINARY Knower-level variable from ordinal measure in data file
GiveN_subset$GN_sm_2cat <- factor(ifelse(GiveN_subset$GN_Small_Ceiling._FinalSet == 6, "CP-knower", "NOT CP-knower"), levels = c("NOT CP-knower", "CP-knower"))

##CREATE PERCENT correct Give-N LARGE measure (second analysis) - drawn from data in 5 separate columns 
GiveN_subset <-  dplyr::mutate(GiveN_subset, t7_corr = ifelse(is.na(GiveN_subset$Target7_FinalSet_Correct.), GiveN_subset$Target7_1st_Response_Fish_Correct., GiveN_subset$Target7_FinalSet_Correct.))
GiveN_subset <-  dplyr::mutate(GiveN_subset, t9_corr = ifelse(is.na(GiveN_subset$Target9_FinalSet_Correct.), GiveN_subset$Target9_1st_Response_Fish_Correct., GiveN_subset$Target9_FinalSet_Correct.))
GiveN_subset <-  dplyr::mutate(GiveN_subset, t10_corr = ifelse(is.na(GiveN_subset$Target10_FinalSet_Correct.), GiveN_subset$Target10_1st_Response_Fish_Correct., GiveN_subset$Target10_FinalSet_Correct.))
GiveN_subset <-  dplyr::mutate(GiveN_subset, t12_corr = ifelse(is.na(GiveN_subset$Target12_FinalSet_Correct.), GiveN_subset$Target12_1st_Response_Fish_Correct., GiveN_subset$Target12_FinalSet_Correct.))
GiveN_subset <-  dplyr::mutate(GiveN_subset, t16_corr = ifelse(is.na(GiveN_subset$Target16_FinalSet_Correct.), GiveN_subset$Target16_1st_Response_Fish_Correct., GiveN_subset$Target16_FinalSet_Correct.))


#CALCULATE MEAN PROP CORR FOR GIVE-N LARGE (only for kids who got all five trials)

# Code below will appropriately not calculate mean (puts "NA") if any trial from above columns has NA value--THIS IS CORRECT, bc doing a proportion correct out of different numbers of trials for different kids is a no-go
GiveN_subset <-  GiveN_subset %>% rowwise() %>% dplyr::mutate(GN_Lg_propcorr_NA=mean(c(t7_corr, t9_corr, t10_corr, t12_corr, t16_corr)))


#Count/prop. of kids who are CP/NON-CP knowers, by Timing
table(GiveN_subset$GN_sm_2cat, GiveN_subset$Language_Timing)
prop.table(table(GiveN_subset$GN_sm_2cat, GiveN_subset$Language_Timing))

```



## Beeswarn Boxplot for GN-2-category (CP-Knower vs. Non-CP-Knower) w/dot colors by Language Timing & Language Modality

```{r}
summary(GiveN_subset$Language_Timing)
table(GiveN_subset$Language_Timing, GiveN_subset$GN_sm_2cat, useNA = "always")
jitter <- position_jitter(width=0, height=0.075)

#N for code below is: summary() values - NA values from table()
a <- ggplot(data=GiveN_subset, mapping = aes(x=GN_sm_2cat, y=Age)) + geom_boxplot() + geom_dotplot(mapping=aes(fill=Language_Timing, color=Language_Timing), method = "dotdensity", binaxis = "y", position = jitter, stackdir = "center", dotsize=0.5, stackratio = 1.25) + theme(text = element_text(size=11), axis.title.x=element_blank(), legend.position = c(0.8, 0.2))  + ylab("Age (Years)") + scale_fill_manual(name = "Language\nTiming", breaks=c("Later", "Early"), values=c("firebrick3", "grey37"), labels=c("Later (83)", "Early (93)")) + scale_color_manual(name = "Language\nTiming", breaks=c("Later", "Early"), values=c("firebrick3", "grey37"), labels=c("Later (83)", "Early (93)"))

summary(GiveN_subset$Language_Modality)
table(GiveN_subset$Language_Modality, GiveN_subset$GN_sm_2cat, useNA = "always")

#N for code below is: summary() values - NA values from table()
b <- ggplot(data=GiveN_subset, mapping = aes(x=GN_sm_2cat, y=Age)) + geom_boxplot() + geom_dotplot(mapping=aes(fill=Language_Modality, color=Language_Modality), binaxis = "y", position = jitter, stackdir = "center", dotsize=0.5, stackratio = 1.25) + theme(text = element_text(size=10), axis.title.x=element_blank(), legend.position = c(0.8, 0.2))  + ylab("Age (Years)") + scale_fill_manual(name = "Language Modality", values=c("tan2", "forestgreen"), breaks=c("English", "ASL"), labels=c("English (105)", "ASL (71)")) + scale_color_manual(name = "Language Modality", values=c("tan2", "forestgreen"), breaks=c("English", "ASL"), labels=c("English (105)", "ASL (71)"))

ggarrange(a, b, ncol=2, nrow=1, common.legend = FALSE)
```

## Figure 4 - SCATTERPLOTS of proportion correct values on GIVE-N Quantities 7, 9, 10, 12, 16 by Age (years), separated by Language Timing & Language Modality
```{r}
summary(GiveN_subset$Language_Timing)
table(GiveN_subset$Language_Timing, GiveN_subset$GN_Lg_propcorr_NA, useNA = "always")

#N for code below is: summary() values - NA values from table()
ggplot(data=GiveN_subset, aes(x=Age, y=GN_Lg_propcorr_NA, shape=Language_Timing, color=Language_Timing)) + geom_jitter(na.rm=TRUE) + geom_smooth(method=lm, se = FALSE) + labs(x="Age at Test (Years)", y="Give-N Large Proportion Correct") + theme(plot.title = element_text(hjust = 0.5)) + scale_shape_manual(name="Language\nTiming", values=c(17, 16), labels=c("Later (75)", "Early (60)")) + scale_color_manual(name="Language\nTiming", values=c("firebrick3", "deepskyblue3"), labels=c("Later (75)", "Early (60)"))


summary(GiveN_subset$Language_Modality)
table(GiveN_subset$Language_Modality, GiveN_subset$GN_Lg_propcorr_NA, useNA = "always")

#N for code below is: summary() values - NA values from table()
ggplot(data=GiveN_subset, aes(x=Age, y=GN_Lg_propcorr_NA, shape=Language_Modality, color=Language_Modality)) + geom_jitter(na.rm=TRUE) + geom_smooth(method=lm, se = FALSE) + labs(x="Age at Test (Years)", y="Give-N Large Proportion Correct") + theme(plot.title = element_text(hjust = 0.5)) + scale_color_manual(name = "Language\nModality", values=c("tan2", "forestgreen"), labels=c("English (87)", "ASL (48)")) + scale_shape_manual(name = "Language\nModality", values=c(19, 15),  labels=c("English (87)", "ASL (48)"))
```







## THE MODELS 

### Give-N Quantities 1-6, Split into 2 Categories (CP-knowers and Non-CP-knowers) - LOGISTIC REGRESSION

```{r}
null <-glm(formula = GN_sm_2cat ~ 1, data = GiveN_subset, family = "binomial")
summary(null) #resid dev: 210.07   AIC 212.07
j <- glm(formula = GN_sm_2cat ~ SES..8.66. + Language_Modality + Age + Language_Timing, data = GiveN_subset, family = "binomial")
summary(j) #resid dev: 162.72   AIC 172.72 ##Age & Timing sig., Modality & SES not.

exp(coef(j)) #ODDS RATIOS
	#Kids in the early exposed group are 2.23 times more likely to have achieved CP-knower status than kids in the later group

# Same regression as above, with age x Timing interaction instead of additive effects
##    k <- glm(formula = GN_sm_2cat ~ SES + Modality + Age*Timing, data = GiveN_subset, family = "binomial")
##    summary(k) #resid dev: 159.75   AIC 171.75
# Timing * Age interaction not sig (but neither is anything else except age)
```

## MODEL FIT MEASURES

```{r}
with(j, null.deviance - deviance)
with(j, df.null - df.residual)
with(j, pchisq(null.deviance - deviance, df.null - df.residual, lower.tail = FALSE))
#Model fits sig. better than a null model, p<.001
```



## MODEL DIAGNOSTICS

```{r}
# from http://www.sthda.com/english/articles/36-classification-methods-essentials/148-logistic-regression-assumptions-and-diagnostics-in-r/

#Step 1: checking that continuous predictros ARE in fact linearly associated with logit of the outcome

# Select only numeric predictors
continuous_pred <- GiveN_subset %>%
  dplyr::select(c("Age", "SES..8.66.")) 
predictors <- colnames(continuous_pred)

#get the predicted probabilities for EACH KID in the whole model (based on their specific values of each predictor)
probabilities <- predict(j, type = "response")

# Bind the logit and tidying the data for plot
continuous_pred <- continuous_pred %>%
  mutate(logit = log(probabilities/(1-probabilities))) %>%
  gather(key = "predictors", value = "predictor.value", -logit)

#scatterplots!
ggplot(continuous_pred, aes(logit, predictor.value))+
  geom_point(size = 0.5, alpha = 0.5) +
  geom_smooth(method = "loess") + 
  theme_bw() + 
  facet_wrap(~predictors, scales = "free_y")



# Step 2: checking for "INFLUENTIAL VALUES" - extreme data points
plot(j, which = 4, id.n = 3)

# Extract model results
j.data <- augment(j) %>% 
  mutate(index = 1:n()) 

j.data %>% top_n(3, .cooksd)

ggplot(j.data, aes(index, .std.resid)) + 
  geom_point(aes(color = GN_sm_2cat), alpha = .5) +
  theme_bw()

j.data %>% 
  filter(abs(.std.resid) > 3)
#NO INFLUENTIAL OBSERVATIONS IN DATA


##Step 3: MULTICOLLINEARITY?
car::vif(j)  ##None! (bc values not >5 or 10)


# WOHOOO WE HAVE A GOOD MODEL, FOLKS!

#output model
stargazer(j, title = "Logistic Regression Results", align=TRUE, dep.var.labels=c("CP-Knower Status"), covariate.labels=c("Socioeconomic Status (SES)", "Language Modality (ASL)", "Age (Years)", "Timing of Language Exposure (Early)"), single.row=TRUE, keep.stat=c("n", "aic"),  out= "GN_CPknowerbinary_SES_Modality_Age_Timing_200514.htm")
```



## GIVE-N QUANTITIES 7, 9, 10, 12, 16
Both linear and tobit models had poor measures of model fit and poor residual plots, so using an ordinal model


### ORDINAL MODEL

```{r}
GN_lg_noNA <- subset(GiveN_subset, GiveN_subset$GN_Lg_propcorr_NA!='')
View(GN_lg_noNA)

GN_lg_noNA$GN_lg_factor <- as.factor(as.character(GN_lg_noNA$GN_Lg_propcorr_NA))
GN_lg_noNA$GN_lg_factor

GN_lg_null <- polr(formula = GN_lg_factor ~ 1, data = GN_lg_noNA,  method = "probit", Hess = TRUE) 
summary(GN_lg_null) #res dev. 410.7124  AIC: 420.7124

GN_lg_polr <- polr(formula = GN_lg_factor ~ SES..8.66. + Language_Modality + Age + Language_Timing, data = GN_lg_noNA,  method = "probit", Hess = TRUE) 
summary(GN_lg_polr)  # res dev 337.9581   AIC: 355.9581


lipsitz.test(GN_lg_polr) #ns
logitgof(GN_lg_noNA$GN_lg_factor, fitted(GN_lg_polr)) #ns
pulkrob.chisq(GN_lg_polr, c("Language_Modality", "Language_Timing")) #ns
```



## checking RESIDUALS
```{r}
sres <- resids(GN_lg_polr)

p1 <- autoplot(sres, what = "qq", distribution = qnorm)
p2 <- autoplot(sres, what = "covariate", x = GN_lg_noNA$SES..8.66., xlab = "SES") #bit of bunching at higher values but we know SES is skewed (also not variable of interest in this model)
p3 <- autoplot(sres, what = "covariate", x = GN_lg_noNA$Language_Modality, xlab = "Language Modality") #good
p4 <- autoplot(sres, what = "covariate", x = GN_lg_noNA$Age, xlab = "Age") #good
p5 <- autoplot(sres, what = "covariate", x = GN_lg_noNA$Language_Timing, xlab = "Language Timing") #good
p6 <- autoplot(sres, what = "covariate", x = GN_lg_noNA$GN_lg_factor, xlab = "Large Number Knowledge") # shows that surrogate residuals underestimated for lower values of Give-N large prop corr (0, 0.2, 0.4, 0.6) and overestimated for  (not sure how much to make of this)
grid.arrange(p1, p2,p3, p4, p5, p6, ncol = 3)  # code that produces Figure 2 in Greenwell et al ppr


## Good fitting model!

stargazer(GN_lg_polr, title = "Ordinal Logistic Regression Results", align=TRUE, dep.var.labels=c("Give-N Large Proportion Correct"), covariate.labels=c("Socioeconomic Status (SES)", "Language Modality (ASL)", "Age (Years)", "Timing of Language Exposure (Later)"), single.row=TRUE, keep.stat=c("n", "aic"),  out= "GN_lgordinal_SES_Modality_Age_Timing_probit_200514.htm")
```

